---
title: "Effects of Weather events (in US) on the Population and Economy"
author: "Vineet"
date: "`r format(Sys.Date(), '%d %B, %Y')`"
output: 
 html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height = 8, fig.width = 12)
```

## Synopsis

In the report, we explore the U.S. National Oceanic and Atmospheric
Administration's (NOAA) storm database. This database tracks
characteristics of major storms and weather events in the United States,
including when and where they occur, as well as estimates of any
fatalities, injuries, and property damage. We will analyze the data to
answer the following questions:

* Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
* Across the United States, which types of events have the greatest economic consequences? 

Our analysis concludes that: 

* Tornadoes are the most harmful for population health
* Floods are the most harmful overall while Hurricanes/Typhoons are the most devastating per event

## Data processing

Loading libraries

```{r}
library(dplyr)
library(ggplot2)
library(scales)
library(gridExtra)
```

Downloading and Reading the file

```{r cache=TRUE}

file_storm <- "./data/repdata_data_StormData.csv.bz2"
if(!file.exists(file_storm)) {
    url_storm <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
    download.file(url = url_storm, destfile = file_storm)
}

df_storm <- read.csv(bzfile(file_storm))
```

After loading the data, we'll check the first few rows and the structure
of the data

```{r}
head(df_storm)
```

For the purpose of our analysis, we will select only the columns that
are relevant to us 
    1. Columns FATALITIES & INJURIES tell us about the effect of storm and weather events on the population health 
    2. The column sets of PROPDMG, PROPDMGEXP & CROPDMG, CROPDMGEXP tell us about the economic damage of storm and weather events by estimating the approximate loss (in dollars) of Property and Crops

```{r}
colsToSelect <- c("BGN_DATE","EVTYPE","FATALITIES","INJURIES","PROPDMG","PROPDMGEXP","CROPDMG","CROPDMGEXP")

df_storm_short <- df_storm[, colsToSelect]
head(df_storm_short)

```

Checking unique values in PROPDMGEXP and CROPDMGEXP to understand the
different types of characters used to indicate the multiplication factor
to be applied to PROPDMG and CROPDMG respectively

```{r}
unique(df_storm_short$PROPDMGEXP)
unique(df_storm_short$CROPDMGEXP)
```

We can understand the following mappings from above: 

* h/H = hundred /'00
* k/K = thousand / '000 
* m/M = million / '000,000
* b/B = billion / '000,000,000

Adding a column to transform the character symbols to corresponding
multiplication factor

```{r}

df_storm_short <- df_storm_short %>% 
    mutate(PROPDMGEXP2 = case_when(
        (PROPDMGEXP == "h" | PROPDMGEXP == "H") ~ 100,
        (PROPDMGEXP == "k" | PROPDMGEXP == "K") ~ 1000,
        (PROPDMGEXP == "m" | PROPDMGEXP == "M") ~ 1000000,
        (PROPDMGEXP == "b" | PROPDMGEXP == "B") ~ 1000000000,
        PROPDMGEXP == "" ~ 1
    ),
    CROPDMGEXP2 = case_when(
        (CROPDMGEXP == "h" | CROPDMGEXP == "H") ~ 100,
        (CROPDMGEXP == "k" | CROPDMGEXP == "K") ~ 1000,
        (CROPDMGEXP == "m" | CROPDMGEXP == "M") ~ 1000000,
        (CROPDMGEXP == "b" | CROPDMGEXP == "B") ~ 1000000000,
        CROPDMGEXP == "" ~ 1
    ))
       
```

For the rest of the characters, we are NOT able to interpret the
meaning. We will check how much data is there, apart from the characters
we identified, to understand the impact this data will have on our
analysis

```{r}
table(df_storm_short$PROPDMGEXP2, useNA = "ifany")
table(df_storm_short$CROPDMGEXP2, useNA = "ifany")
```

We can see here that the NA values OR the no of observations for which
we couldn't interpret the characters, account for a very small
percentage of the total no of observations. We will ignore the
transformation for these observations.

calculating the economic damage value with the derived multiplication
factor

```{r}

df_storm_short <- df_storm_short %>% 
    mutate(PROPDMGVAL = PROPDMG*PROPDMGEXP2,
           CROPDMGVAL = CROPDMG*CROPDMGEXP2)
```

## Results

#### Question 1: Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

```{r}

#fatalities2 <- tapply(df_storm_short$FATALITIES, df_storm_short$EVTYPE, sum)
#fatalities2 <- sort(fatalities2, decreasing = TRUE)
#fatalities2 <- fatalities2[1:10]

fatalities <- aggregate(FATALITIES ~ EVTYPE, df_storm_short, sum)
fatalities <- fatalities[order(-fatalities$FATALITIES), ]
fatalities <- fatalities[1:10,]
fatalities$EVTYPE <- factor(fatalities$EVTYPE, levels = fatalities$EVTYPE)

p1 <- ggplot(fatalities, aes(x = EVTYPE, y = FATALITIES)) + geom_bar(stat = "identity") + xlab("Event Type") + ylab("No of Fatalities") + ggtitle("Most Fatalities per Weather Event") + theme(axis.text.x = element_text(angle = 45, hjust = 1))

#injuries2 <- tapply(df_storm_short$INJURIES, df_storm_short$EVTYPE, sum)
#injuries2 <- sort(injuries2, decreasing = TRUE)
#injuries2 <- injuries2[1:10]

injuries <- aggregate(INJURIES ~ EVTYPE, df_storm_short, sum)
injuries <- injuries[order(-injuries$INJURIES),]
injuries <- injuries[1:10,]
injuries$EVTYPE <- factor(injuries$EVTYPE, levels = injuries$EVTYPE)

p2 <- ggplot(injuries, aes(x = EVTYPE, y = INJURIES)) + geom_bar(stat = "identity") + xlab("Event Type")+ ylab("No of Injuries") + ggtitle("Most Injuries per Weather Event") + theme(axis.text.x = element_text(angle = 45, hjust = 1))

grid.arrange(p1,p2, ncol = 2)

```

#### Answer 1 : Tornado is the most harmful weather event with respect to population health

#### Question 2: Across the United States, which types of events have the greatest economic consequences?

```{r}
damages <- aggregate(PROPDMGVAL + CROPDMGVAL ~ EVTYPE, df_storm_short, sum)
names(damages) <- c("EVTYPE", "TOTDMG")
damages <- damages[order(-damages$TOTDMG), ]
damages <- damages[1:10, ]
damages$EVTYPE <- factor(damages$EVTYPE, levels = damages$EVTYPE)

p3 <- ggplot(damages, aes(x = EVTYPE, y = TOTDMG)) + geom_bar(stat = "identity") + xlab("Event Type") +  ylab("Damages $") + ggtitle("Most Damgages (in USD) per Weather Event") + theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7)) + scale_y_continuous(labels = label_number(suffix = " Bn", scale = 1e-09))

damages_nObs <- df_storm_short %>% 
    count(EVTYPE)
damages_nObs <- damages_nObs[damages_nObs$EVTYPE %in% damages[,1],]
damages_nObs$EVTYPE <- factor(damages_nObs$EVTYPE, levels = damages$EVTYPE)

p4<- ggplot(damages_nObs, aes(x = EVTYPE, y = n)) + geom_bar(stat = "identity") + geom_text(aes(label = n), vjust = -0.3, size = 3) + xlab("Event Type") + ylab("No of Events") + ggtitle("No of Events per Event Type") + theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7)) + scale_y_continuous(labels = label_number(suffix = "K", scale = 1e-03))

grid.arrange(p3,p4)
```

#### Answer 2: Flood has had the greatest economic consequence. However, by looking at the corresponding no of events - Hurricanes and Typhoons are the deadliest weather events with greater economic damage per event.
